---
title: "CPT_S 575 Project"
author: "Najarian & Shatanawi"
date: "2022-12-04"
output:
  html_document:
    df_print: paged
---

# Load data and libraries
```{r,echo = TRUE}
library(readxl)
columbia_river_flow_data_2015_2022 <- read_excel("C:/Users/johnn/OneDrive - Washington State University (email.wsu.edu)/Classes/CPT_S 575/Project/Data_science_Project_to john/columbia_river_flow_data_2015-2022.xlsx")
View(columbia_river_flow_data_2015_2022)

library(ggplot2)
library(tidyr)
library(dplyr)
#install.packages("imputeTS")
library(imputeTS)

```

# Clean Data
## Fill Missing Vlaues
```{r,echo = TRUE}
#Inflow
Rocky_Reach_Inflow <- c(na_ma(columbia_river_flow_data_2015_2022$Rocky_Reach_Inflow, k = 1))
Rock_Island_Inflow <- c(na_ma(columbia_river_flow_data_2015_2022$Rock_Island_Inflow, k = 1))

#Discharge
Rocky_Reach_Total_Discharge <- c(na_ma(columbia_river_flow_data_2015_2022$Rocky_Reach_Total_Discharge, k = 1))
Rock_Island_Total_Discharge <- c(na_ma(columbia_river_flow_data_2015_2022$Rock_Island_Total_Discharge, k = 1))

#Headwaters
Rocky_Reach_Headwater_Elevation <- c(na_ma(columbia_river_flow_data_2015_2022$Rocky_Reach_Headwater_Elevation, k = 1))
Rock_Island_Headwater_Elevation <- c(na_ma(columbia_river_flow_data_2015_2022$Rock_Island_Headwater_Elevation, k = 1))

#Time
datetime_utc <- columbia_river_flow_data_2015_2022$datetime_utc
datetime_stamp <- columbia_river_flow_data_2015_2022$datetime_stamp

#Construct clean Matrix
Clean_CR_FlowData <- data.frame(datetime_utc, datetime_stamp, Rocky_Reach_Inflow, Rock_Island_Inflow, Rocky_Reach_Total_Discharge, Rock_Island_Total_Discharge, Rocky_Reach_Headwater_Elevation, Rock_Island_Headwater_Elevation)

#Netflow = inflow-discharge
Clean_CR_FlowData$Rocky_Reach_Netflow <- c(Rocky_Reach_Inflow-Rocky_Reach_Total_Discharge)
Clean_CR_FlowData$Rock_Island_Netflow <- c(Rock_Island_Inflow-Rock_Island_Total_Discharge)

```

## Removing Feb. 29th
```{r,echo = TRUE}
Clean_CR_FlowData <- separate(Clean_CR_FlowData, col=datetime_stamp, into=c('year', 'month', 'day'), sep='-')
Clean_CR_FlowData <- separate(Clean_CR_FlowData, col=day, into=c('day', 'time'), sep=' ')
Clean_CR_FlowData = filter(Clean_CR_FlowData,!month =='02' & !day == '29')
```

## Splitting up data by year
```{r,echo = TRUE}
data_2015 <- filter(Clean_CR_FlowData, grepl('2015',datetime_utc)) #missing some from JAN
data_2016 <- filter(Clean_CR_FlowData, grepl('2016',datetime_utc)) #leap year
data_2017 <- filter(Clean_CR_FlowData, grepl('2017',datetime_utc)) #normal year 
data_2018 <- filter(Clean_CR_FlowData, grepl('2018',datetime_utc)) #normal year
data_2019 <- filter(Clean_CR_FlowData, grepl('2019',datetime_utc)) #normal year
data_2020 <- filter(Clean_CR_FlowData, grepl('2020',datetime_utc)) #leap year
```
2015 - Incomplete set 
2016 - Training set 
2017 - Training set 
2018 - Training set 
2019 - Training set 
2020 - Test set

# Modeling Inflow
## Identifying what degree of polynomial is best suited for Time vs Inflow
```{r,echo = TRUE}
degree <- 10
mse = data.frame(data=NA,nrow=degree,ncol=4)

for (i in 1:degree){
        #2016 vs 2020
        fit.train = lm(Rocky_Reach_Inflow ~ poly(c(1:93888),i), data=data_2016)
        fit.test = predict(fit.train)
        mse[i,1] = mean((fit.test-data_2020$Rocky_Reach_Inflow)^2) 
        
        #2017 vs 2020
        fit.train = lm(Rocky_Reach_Inflow ~ poly(c(1:93888),i), data=data_2017)
        fit.test = predict(fit.train)
        mse[i,2] = mean((fit.test-data_2020$Rocky_Reach_Inflow)^2)  
        
        #2018 vs 2020
        fit.train = lm(Rocky_Reach_Inflow ~ poly(c(1:93888),i), data=data_2018)
        fit.test = predict(fit.train)
        mse[i,3] = mean((fit.test-data_2020$Rocky_Reach_Inflow)^2) 
        
        #2019 vs 2020
        fit.train = lm(Rocky_Reach_Inflow ~ poly(c(1:93888),i), data=data_2019)
        fit.test = predict(fit.train)
        mse[i,4] = mean((fit.test-data_2020$Rocky_Reach_Inflow)^2) 
}

#Rename mse dataframe column names
colnames(mse)[1] = "MSE_2016" 
colnames(mse)[2] = "MSE_2017" 
colnames(mse)[3] = "MSE_2018" 
colnames(mse)[4] = "MSE_2019" 

#Adding a degree polynomial column 
mse$degree <- c(1:degree)

ggplot(mse, aes(x=degree)) +  
  geom_line(aes(y = MSE_2016, col = "2016")) + 
  geom_line(aes(y = MSE_2017, col = "2017")) + 
  geom_line(aes(y = MSE_2018, col = "2018")) + 
  geom_line(aes(y = MSE_2019, col = "2019")) +
    scale_x_continuous(breaks = c(1:degree))+
  xlab("Degree of Polynomial")+
  ylab("Mean Squared Error")+
  guides(color = guide_legend(title = "Year"))+
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
mse
```

## Plotting 5th Degree Polynomials for Time vs Inflow
```{r,echo = TRUE}
Time_Inflow_Model_2016 = lm(Rocky_Reach_Inflow ~ poly(c(1:93888),5), data=data_2016)
ggplot(data_2016, aes(c(1:93888), Rocky_Reach_Inflow)) +
  geom_point(size = .01, color = 1, shape = 1)+
  ggtitle("Time vs Inflow - Rocky Reach Dam - 2016")+
  xlab("Time")+
  ylab("Inflow (Kcfs)")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  stat_smooth(method = "lm",
              formula = y ~ poly(x, 5),
              se = FALSE)
Time_Inflow_Model_2017 = lm(Rocky_Reach_Inflow ~ poly(c(1:93888),5), data=data_2017)
ggplot(data_2017, aes(c(1:93888), Rocky_Reach_Inflow)) +
  geom_point(size = .01, color = 1, shape = 1)+
  ggtitle("Time vs Inflow - Rocky Reach Dam - 2017")+
  xlab("Time")+
  ylab("Inflow (Kcfs)")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  stat_smooth(method = "lm",
              formula = y ~ poly(x, 5),
              se = FALSE)
Time_Inflow_Model_2018 = lm(Rocky_Reach_Inflow ~ poly(c(1:93888),5), data=data_2018)
ggplot(data_2018, aes(c(1:93888), Rocky_Reach_Inflow)) +
  geom_point(size = .01, color = 1, shape = 1)+
  ggtitle("Time vs Inflow - Rocky Reach Dam - 2018")+
  xlab("Time")+
  ylab("Inflow (Kcfs)")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  stat_smooth(method = "lm",
              formula = y ~ poly(x, 5),
              se = FALSE)
Time_Inflow_Model_2019 = lm(Rocky_Reach_Inflow ~ poly(c(1:93888),5), data=data_2019)
ggplot(data_2019, aes(c(1:93888), Rocky_Reach_Inflow)) +
  geom_point(size = .01, color = 1, shape = 1)+
  ggtitle("Time vs Inflow - Rocky Reach Dam - 2019")+
  xlab("Time")+
  ylab("Inflow (Kcfs)")+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  stat_smooth(method = "lm",
              formula = y ~ poly(x, 5),
              se = FALSE)
```

# Modeling Headwater
## Identifying what degree of polynomial is best suited for Inflow vs Headwater
```{r,echo = TRUE}
degree <- 10
mse = data.frame(data=NA,nrow=degree,ncol=4)

for (i in 1:degree){
        #2016 vs 2020
        fit.train = lm(Rocky_Reach_Headwater_Elevation ~ poly(Rocky_Reach_Inflow,i), data=data_2016)
        fit.test = predict(fit.train, newdata=data_2020)
        mse[i,1] = mean((fit.test-data_2020$Rocky_Reach_Headwater_Elevation)^2) 
        
        #2017 vs 2020
        fit.train = lm(Rocky_Reach_Headwater_Elevation ~ poly(Rocky_Reach_Inflow,i), data=data_2017)
        fit.test = predict(fit.train, newdata=data_2020)
        mse[i,2] = mean((fit.test-data_2020$Rocky_Reach_Headwater_Elevation)^2)  
        
        #2018 vs 2020
        fit.train = lm(Rocky_Reach_Headwater_Elevation ~ poly(Rocky_Reach_Inflow,i), data=data_2018)
        fit.test = predict(fit.train, newdata=data_2020)
        mse[i,3] = mean((fit.test-data_2020$Rocky_Reach_Headwater_Elevation)^2) 
        
        #2019 vs 2020
        fit.train = lm(Rocky_Reach_Headwater_Elevation ~ poly(Rocky_Reach_Inflow,i), data=data_2019)
        fit.test = predict(fit.train, newdata=data_2020)
        mse[i,4] = mean((fit.test-data_2020$Rocky_Reach_Headwater_Elevation)^2) 
}

#Rename mse dataframe column names
colnames(mse)[1] = "MSE_2016" 
colnames(mse)[2] = "MSE_2017" 
colnames(mse)[3] = "MSE_2018" 
colnames(mse)[4] = "MSE_2019" 

#Adding a degree polynomial column 
mse$degree <- c(1:degree)

ggplot(mse, aes(x=degree)) +  
  geom_line(aes(y = MSE_2016, col = "2016")) + 
  geom_line(aes(y = MSE_2017, col = "2017")) + 
  geom_line(aes(y = MSE_2018, col = "2018")) + 
  geom_line(aes(y = MSE_2019, col = "2019")) +
  ylim(0.2,.7) +
  xlim(0,10)+ 
  scale_x_continuous(breaks = c(1:degree))+
  xlab("Degree of Polynomial")+
  ylab("Mean Squared Error")+
  guides(color = guide_legend(title = "Year"))

mse
```

## Plot Inflow vs Headwater with linear regression
```{r,echo = TRUE}
Inflow_Headwater_2016 <- ggplot(data_2016, aes(x=Rocky_Reach_Inflow,y = Rocky_Reach_Headwater_Elevation)) + 
  geom_point(size = .01, color = 1, shape = 1) +
  ylim(700, 710)+
  xlim(0, 250)+
  geom_smooth(method='lm')+
  ggtitle("Inflow vs Headwater- Rocky Reach Dam - 2016")+
  xlab("Inflow (kcfs)")+
  ylab("Headwater (ft)")
Inflow_Headwater_2016

Inflow_Headwater_2017 <- ggplot(data_2017, aes(x=Rocky_Reach_Inflow,y = Rocky_Reach_Headwater_Elevation)) + 
  geom_point(size = .01, color = 1, shape = 1) +
  ylim(700, 710)+
  xlim(0, 250)+
  geom_smooth(method='lm')+
  ggtitle("Inflow vs Headwater- Rocky Reach Dam - 2017")+
  xlab("Inflow (kcfs)")+
  ylab("Headwater (ft)")
Inflow_Headwater_2017

Inflow_Headwater_2018 <- ggplot(data_2018, aes(x=Rocky_Reach_Inflow,y = Rocky_Reach_Headwater_Elevation)) + 
  geom_point(size = .01, color = 1, shape = 1) +
  ylim(700, 710)+
  xlim(0, 250)+
  geom_smooth(method='lm')+
  ggtitle("Inflow vs Headwater- Rocky Reach Dam - 2018")+
  xlab("Inflow (kcfs)")+
  ylab("Headwater (ft)")
Inflow_Headwater_2018

Inflow_Headwater_2019 <- ggplot(data_2019, aes(x=Rocky_Reach_Inflow,y = Rocky_Reach_Headwater_Elevation)) + 
  geom_point(size = .01, color = 1, shape = 1) +
  ylim(700, 710)+
  xlim(0, 250)+
  geom_smooth(method='lm')+
  ggtitle("Inflow vs Headwater- Rocky Reach Dam - 2019")+
  xlab("Inflow (kcfs)")+
  ylab("Headwater (ft)")
Inflow_Headwater_2019
```

# Modeling Discharge
## Identifying what degree of polynomial is best suited for Headwater vs Discharge
```{r,echo = TRUE}
degree <- 10
mse = data.frame(data=NA,nrow=degree,ncol=4)

for (i in 1:degree){
        #2016 vs 2020
        fit.train = lm(Rocky_Reach_Total_Discharge ~ poly(Rocky_Reach_Headwater_Elevation,i), data=data_2016)
        fit.test = predict(fit.train, newdata=data_2020)
        mse[i,1] = mean((fit.test-data_2020$Rocky_Reach_Total_Discharge)^2) 
        
        #2017 vs 2020
        fit.train = lm(Rocky_Reach_Total_Discharge ~ poly(Rocky_Reach_Headwater_Elevation,i), data=data_2017)
        fit.test = predict(fit.train, newdata=data_2020)
        mse[i,2] = mean((fit.test-data_2020$Rocky_Reach_Total_Discharge)^2)   
        
        #2018 vs 2020
       fit.train = lm(Rocky_Reach_Total_Discharge ~ poly(Rocky_Reach_Headwater_Elevation,i), data=data_2018)
        fit.test = predict(fit.train, newdata=data_2020)
        mse[i,3] = mean((fit.test-data_2020$Rocky_Reach_Total_Discharge)^2) 
        
        #2019 vs 2020
        fit.train = lm(Rocky_Reach_Total_Discharge ~ poly(Rocky_Reach_Headwater_Elevation,i), data=data_2019)
        fit.test = predict(fit.train, newdata=data_2020)
        mse[i,4] = mean((fit.test-data_2020$Rocky_Reach_Total_Discharge)^2) 
}

#Rename mse dataframe column names
colnames(mse)[1] = "MSE_2016" 
colnames(mse)[2] = "MSE_2017" 
colnames(mse)[3] = "MSE_2018" 
colnames(mse)[4] = "MSE_2019" 

#Adding a degree polynomial column 
mse$degree <- c(1:degree)

ggplot(mse, aes(x=degree)) +  
  geom_line(aes(y = MSE_2016, col = "2016")) + 
  geom_line(aes(y = MSE_2017, col = "2017")) + 
  geom_line(aes(y = MSE_2018, col = "2018")) + 
  geom_line(aes(y = MSE_2019, col = "2019")) +
    scale_x_continuous(breaks = c(1:degree))+
  xlab("Degree of Polynomial")+
  ylab("Mean Squared Error")+
  guides(color = guide_legend(title = "Year"))+
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
mse
```

## Plotting 5th Degree Polynomials for Headwater vs Discharge
```{r,echo = TRUE}
ggplot(data_2016, aes(x = Rocky_Reach_Headwater_Elevation,y = Rocky_Reach_Total_Discharge))+
  geom_point(size = .01, color = 1, shape = 1) +
  ggtitle("Headwater vs Discharge - Rocky Reach Dam - 2016")+
  xlab("Headwater")+
  ylab("Discharge")+
  ylim(0, 250)+
  xlim(700, 710)+
  stat_smooth(method = "lm",
              formula = y ~ poly(x, 1),
              se = FALSE)

ggplot(data_2017, aes(x = Rocky_Reach_Headwater_Elevation,y = Rocky_Reach_Total_Discharge))+
  geom_point(size = .01, color = 1, shape = 1) +
  ggtitle("Headwater vs Discharge - Rocky Reach Dam - 2017")+
  xlab("Headwater")+
  ylab("Discharge")+
  ylim(0, 250)+
  xlim(700, 710)+
  stat_smooth(method = "lm",
              formula = y ~ poly(x, 1),
              se = FALSE)

ggplot(data_2018, aes(x = Rocky_Reach_Headwater_Elevation,y = Rocky_Reach_Total_Discharge))+
  geom_point(size = .01, color = 1, shape = 1) +
  ggtitle("Headwater vs Discharge - Rocky Reach Dam - 2018")+
  xlab("Headwater")+
  ylab("Discharge")+
  ylim(0, 250)+
  xlim(700, 710)+
  stat_smooth(method = "lm",
              formula = y ~ poly(x, 1),
              se = FALSE)

ggplot(data_2019, aes(x = Rocky_Reach_Headwater_Elevation,y = Rocky_Reach_Total_Discharge))+
  geom_point(size = .01, color = 1, shape = 1) +
  ggtitle("Headwater vs Discharge - Rocky Reach Dam - 2019")+
  xlab("Headwater")+
  ylab("Discharge")+
  ylim(0, 250)+
  xlim(700, 710)+
  stat_smooth(method = "lm",
              formula = y ~ poly(x, 1),
              se = FALSE)
```

## Solving for 2020 Discharge 
```{r,echo = TRUE}
#2016 Modeled Discharges
Inflow_Model_2016 <- lm(Rocky_Reach_Inflow ~ poly(c(1:93888),5), data=data_2020)
Modeled_Inflows_2016 <- data.frame(predict(Inflow_Model_2016))
colnames(Modeled_Inflows_2016)[1] = "Modeled_Inflows"
Headwater_Model_2016 <- lm(data_2016$Rocky_Reach_Headwater_Elevation ~ Modeled_Inflows_2016$Modeled_Inflows)
Modeled_Headwaters_2016 <- data.frame(predict(Headwater_Model_2016))
colnames(Modeled_Headwaters_2016)[1] = "Modeled_Headwaters"
Discharge_Model_2016 <- lm(data_2016$Rocky_Reach_Total_Discharge ~ Modeled_Headwaters_2016$Modeled_Headwaters)
Modeled_Discharges_2016 <- data.frame(predict(Discharge_Model_2016))
colnames(Modeled_Discharges_2016)[1] = "Modeled_Discharges"

#2017 Modeled Discharges
Inflow_Model_2017 <- lm(Rocky_Reach_Inflow ~ poly(c(1:93888),5), data=data_2020)
Modeled_Inflows_2017 <- data.frame(predict(Inflow_Model_2017))
colnames(Modeled_Inflows_2017)[1] = "Modeled_Inflows"
Headwater_Model_2017 <- lm(data_2017$Rocky_Reach_Headwater_Elevation ~ Modeled_Inflows_2017$Modeled_Inflows)
Modeled_Headwaters_2017 <- data.frame(predict(Headwater_Model_2017))
colnames(Modeled_Headwaters_2017)[1] = "Modeled_Headwaters"
Discharge_Model_2017 <- lm(data_2017$Rocky_Reach_Total_Discharge ~ Modeled_Headwaters_2017$Modeled_Headwaters)
Modeled_Discharges_2017 <- data.frame(predict(Discharge_Model_2017))
colnames(Modeled_Discharges_2017)[1] = "Modeled_Discharges"

#2018 Modeled Discharges
Inflow_Model_2018 <- lm(Rocky_Reach_Inflow ~ poly(c(1:93888),5), data=data_2020)
Modeled_Inflows_2018 <- data.frame(predict(Inflow_Model_2018))
colnames(Modeled_Inflows_2018)[1] = "Modeled_Inflows"
Headwater_Model_2018 <- lm(data_2018$Rocky_Reach_Headwater_Elevation ~ Modeled_Inflows_2018$Modeled_Inflows)
Modeled_Headwaters_2018 <- data.frame(predict(Headwater_Model_2018))
colnames(Modeled_Headwaters_2018)[1] = "Modeled_Headwaters"
Discharge_Model_2018 <- lm(data_2018$Rocky_Reach_Total_Discharge ~ Modeled_Headwaters_2018$Modeled_Headwaters)
Modeled_Discharges_2018 <- data.frame(predict(Discharge_Model_2018))
colnames(Modeled_Discharges_2018)[1] = "Modeled_Discharges"

#2019 Modeled Discharges
Inflow_Model_2019 <- lm(Rocky_Reach_Inflow ~ poly(c(1:93888),5), data=data_2020)
Modeled_Inflows_2019 <- data.frame(predict(Inflow_Model_2019))
colnames(Modeled_Inflows_2019)[1] = "Modeled_Inflows"
Headwater_Model_2019 <- lm(data_2019$Rocky_Reach_Headwater_Elevation ~ Modeled_Inflows_2019$Modeled_Inflows)
Modeled_Headwaters_2019 <- data.frame(predict(Headwater_Model_2019))
colnames(Modeled_Headwaters_2019)[1] = "Modeled_Headwaters"
Discharge_Model_2019 <- lm(data_2019$Rocky_Reach_Total_Discharge ~ Modeled_Headwaters_2019$Modeled_Headwaters)
Modeled_Discharges_2019 <- data.frame(predict(Discharge_Model_2019))
colnames(Modeled_Discharges_2019)[1] = "Modeled_Discharges"

mse_2016 <- mean((Modeled_Discharges_2016$Modeled_Discharges-data_2020$Rocky_Reach_Total_Discharge)^2)
mse_2017 <- mean((Modeled_Discharges_2017$Modeled_Discharges-data_2020$Rocky_Reach_Total_Discharge)^2)
mse_2018 <- mean((Modeled_Discharges_2018$Modeled_Discharges-data_2020$Rocky_Reach_Total_Discharge)^2)
mse_2019 <- mean((Modeled_Discharges_2019$Modeled_Discharges-data_2020$Rocky_Reach_Total_Discharge)^2)

Modeled_Discharges <- data.frame(data_2020$Rocky_Reach_Total_Discharge, Modeled_Discharges_2016$Modeled_Discharges, Modeled_Discharges_2017$Modeled_Discharges, Modeled_Discharges_2018$Modeled_Discharges, Modeled_Discharges_2019$Modeled_Discharges)

colnames(Modeled_Discharges)[1] = "Actual_2020"
colnames(Modeled_Discharges)[2] = "Predicted_2020_from_2016"
colnames(Modeled_Discharges)[3] = "Predicted_2020_from_2017"
colnames(Modeled_Discharges)[4] = "Predicted_2020_from_2018"
colnames(Modeled_Discharges)[5] = "Predicted_2020_from_2019"

ggplot(Modeled_Discharges, aes(x=c(1:93888))) +  
  geom_point(aes(y = Actual_2020, col = "2020"),size = .01, color = "yellow", shape = 1) + 
  geom_line(aes(y = Predicted_2020_from_2016, col = "Predicted from 2016 Model")) + 
  geom_line(aes(y = Predicted_2020_from_2017, col = "Predicted from 2017 Model")) + 
  geom_line(aes(y = Predicted_2020_from_2018, col = "Predicted from 2018 Model")) + 
  geom_line(aes(y = Predicted_2020_from_2019, col = "Predicted from 2019 Model")) +
  xlab("Time (Jan-Dec)")+
  ylab("Discharge (kcfs)")+
  guides(color = guide_legend(title = ""))+
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```